---
title: "Bellabeats Market Research Report"
author: "Jerard"
date: "05/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(lubridate)
library(data.table)
library(plyr)
library(zoo)
library(lookup)
library(colorBlindness)
setwd("C:/Users/pc/Documents/GitHub/DataAnalyticsR/DataAnalyticsR")
load("C:/Users/PC/Documents/GitHub/DataAnalyticsR/DataAnalyticsR/.RData")
```

# Bellabeats Market Research Report

## Project Summary 

The objective was to report on how other non-Bellabeats products is being used through a data driven analysis. Based on these results a product from Bellabeats product line was selected to assist the marketing team with their marketing strategy. The data source was crowd sourced from FitBit Fitness tracker data and downloaded from Kaggle. Unfortunately, multiple datasets were incomplete and critical data was missing i.e. sex, age, weight, fitness level and height. Taking these limitations into consideration this project focused on 'daily data' as these datasets were complete and allowed for the most comprehensive analysis to determine how the product was used. 
Based on the analysis the fitness tracker was used primarily as an activity tracker, with all participants wearing the trackers during the day. Although, sleep data was incomplete it allowed for insights into the participants sleeping behaviors. The relationships of various activity levels was analyzed against calories used. A strong positive correlation was found between very active minutes and calories used, while a negative correlation was observed between sedentary minutes and calories used. According to the CDC it is recommended that a person should exercise for a minimum of 22 minutes per day, with the intensity being moderate to vigorous. Looking at the data alone it appears that the participants meet this requirement, however, on closer observation it was noted that the intensity of this exercise is light.
The recommendation from the data is that Bellabeats marketing team should focus on marketing the 'Bellabeats Time' and 'Bellabeats Leaf' as both of these products are able to perform activity tracking. Based on the observations it is recommended that the app should include goals and rewards specific to the user eg. if user aims to lose weight the app should award the user each time they perform vigorous activities, as this would correspond to more calories burned. 


## Data Sources for Analysis

The data source used during this analysis was FitBit Fitness Tracker Data (CC0: Public Domain, dataset made available through Mobius). The dataset contains personal fitness information from thirty (30) FitBit users. Thirty eligible FitBit users consented to the submission of personal tracker data, including     minute-level output for physical activity, heart rate, and sleep monitoring. It includes information about daily activity, steps, and heart rate that can be used to identify users’ habits. 

The metadata provided on Kaggle can be found on https://www.kaggle.com/arashnic/fitbit/metadata. To verify the source I navigated to https://zenodo.org/record/53894#.X9oeh3Uzaao and noticed the data was captured on another application named 'Amazon Mechanical Turk' as part of a survey. Additional verification was not possible. Unfortunately, the original source was unavailable and the eligibility criteria was not disclosed. 

### Critical Analysis of the data source

Unfortunately, the data provided and the information it contained was unreliable. Due to the unavailability of metadata for the original data, which makes it difficult to verify the data's quality. It cannot be established if the data is biased or not as the data source only states that there are thirty eligible Fitbit users. Information such as sex, age, fitness level, weight and height is absent. Furthermore, there are multiple incomplete data tables.  The data appears to be pre-processed as the file name contains 'merged' in the file name (eg. activitydata_merged). The data used is sourced from a Third Party and cannot be validated as the original data is unavailable. There are some critical information missing to perform a more comprehensive analysis i.e. age, sex, height and weight. The data was collected in 2016 from my understanding of the fitness device market there have been multiple improvements over the past five (5) years. The data source is cited.

The reference for the sourced data is as follow:

Furberg, R., Brinton, J., Keating, M., & Ortiz, A. (2016). Crowd-sourced Fitbit datasets 03.12.2016-05.12.2016 [Data set]. Zenodo.     https://doi.org/10.5281/zenodo.53894

Because there is missing information the creditably of the data cannot be verified. This makes it difficult to draw informed conclusions, as the data has glaring weaknesses. Without acknowledging these limitations to the analysis the decisions based on the analysis can be incorrect and cost the company revenue. 


## Preparing the data for analysis

The data sets were downloaded from Kaggle at https://www.kaggle.com/arashnic/fitbit.

The CSV files were imported into R. 

```{r}
# Import data from csv to data frames

activity_daily <- read.csv(file = "dailyActivity_merged.csv", header = TRUE, sep = ",")
heartrate_seconds <- read.csv(file = "heartrate_seconds_merged.csv", header = TRUE, sep = ",")
mets_minute <- read.csv(file = "minuteMETsNarrow_merged.csv", header = TRUE, sep = ",")
sleep_daily <- read.csv(file = "sleepDay_merged.csv", header = TRUE, sep = ",")
sleep_minute <- read.csv(file = "minuteSleep_merged.csv", header = TRUE, sep = ",")
```


The focus was on 'daily data' with the exception of sleep data, heart rate and METs. Viewing the data to ensure that the data has successfully imported. 

```{r}
# View Data to ensure import was successful. The focus will be on the daily data with the exception of the heart rate and sleep data.
tibble(activity_daily)
tibble(heartrate_seconds)
tibble(mets_minute)
tibble(sleep_daily)
tibble(sleep_minute)
```

With the tibbles I noticed a few things. First, was the use of unconventional column name like 'ActivityDate' rather than just 'Date' or the use of 'Time' when in fact it was 'Date_Time'. To ensure that the headers were all in the same format, the clean_names function was used. 

```{r}
# Cleaning process with janitor

# Based on the results the column names for date and time to be made consistent

activity_daily_cln <- activity_daily %>% 
  rename(c("ActivityDate" = "date")) %>% 
  clean_names()

# Heart rate data column to be changed from time to date time

heartrate_seconds_cln <- heartrate_seconds %>% 
  rename(c("Time" = "date_time")) %>% 
  clean_names()

# METs data column name change from activity minutes to date time 

mets_minute_cln <- mets_minute %>% 
  rename(c("ActivityMinute" = "date_time", "METs" = "mets")) %>% 
  clean_names()

# Sleep data column names to be changed the SleepDay to Date

sleep_daily_cln <- sleep_daily %>% 
  rename(c("SleepDay" = "date")) %>% 
  clean_names()

# Sleep minute column name date to date time

sleep_minute_cln <- sleep_minute %>% 
  rename(c("date" = "date_time")) %>% 
  clean_names()
```

The tabyl function was used to verify the completeness of the data and provided summaries of each data frame.

```{r}
# Produce a table summarizing the table based on id.

tabyl(activity_daily_cln, id)

# The outcome: there was only 21 people that wore the fitness device for 31 days

tabyl(activity_daily_cln, id) %>% 
  summarise(count(n))

# The heart rate was more incomplete that the activity data with only 14 participants 

tabyl(heartrate_seconds_cln,id) %>% 
  summarise(count(n))

# There was 33 observations in total of varying lengths

tabyl(mets_minute_cln, id) %>% 
  summarise(count(n))
  
# There were 24 participants, tracking their sleep a minimum of 1 and a maximum of 32 days.

tabyl(sleep_daily_cln, id) %>% 
  summarise(count(n))

# There were 24 participants, tracking their sleep a minimum of 69 min and a maximum of 15682 min. This is consistent with the observations noted above.

tabyl(sleep_minute_cln, id) %>% 
  summarise(count(n))

```


The next step was to convert the character format for date and time to date formats as the focus was on 'daily data'.		

```{r,warning=FALSE}
# To be able to plot the data the date formats of all the dataframes needs to be changed

# Date is changed for the two dataframes below

activity_daily_cln_v2[[2]] <- as.Date(activity_daily_cln[[2]], "%m/%d/%Y")
tibble(activity_daily_cln_v2)

sleep_daily_cln_v2[[2]] <- as.Date(sleep_daily_cln[[2]], "%m/%d/%Y")
tibble(sleep_daily_cln_v2)

# The dataframes below needs datetime format

heartrate_seconds_cln_v2[[2]] <- as.POSIXct(heartrate_seconds_cln[[2]], format = "%m/%d/%Y %H:%M:%S %p")
tibble(heartrate_seconds_cln_v2)

mets_minute_cln_v2[[2]] <- as.POSIXct(mets_minute_cln[[2]], format = "%m/%d/%Y %H:%M:%S %p")
tibble(mets_minute_cln_v2)

# Separating column for the date time in tables for heart rate and METs

heartrate_seconds_cln_v3 <- separate(heartrate_seconds_cln_v2, col = date_time, c("date","time"), sep = " ")
heartrate_seconds_cln_v3[[2]] <- as.Date(heartrate_seconds_cln_v3[[2]], "%Y-%m-%d")
tibble(mets_minute_cln_v3)

mets_minute_cln_v3 <- separate(mets_minute_cln_v2, col = date_time, c("date","time"), sep = " ")
mets_minute_cln_v3[[2]] <- as.Date(mets_minute_cln_v3[[2]], "%Y-%m-%d")
tibble(mets_minute_cln_v3)

```

Because there was a focus on 'daily data' the heart rate and METs data had to be converted to 'daily data'.

```{r, warning=FALSE}
# Summarize the min and sec data into daily data in order to use for the daily insights.

heartrate_seconds_cln_dt <- data.table(heartrate_seconds_cln_v4)
heartrate_seconds_cln_dt_2 <- heartrate_seconds_cln_dt[,list(mean_hrt = mean(value), max_hrt = max(value), min_hrt = min(value)), by = c("id,date")]
  
  
mets_minute_cln_dt <- data.table(mets_minute_cln_v4)
mets_minute_cln_dt_2 <- mets_minute_cln_dt[,list(mean_mets = mean(mets), max_mets = max(mets), min_mets = min(mets)), by = c("id,date")]
```

Verification to look for duplication's, missing values and unusual values.

```{r, warning=FALSE}
# Looking for any duplication or missing values

# No duplicates 
activity_daily_cln_v2 %>% 
  get_dupes() %>% 
  summarise(count(id))


# Multiple duplicates  
heartrate_seconds_cln_v3 %>% 
  get_dupes() %>% 
  summarise(count(id))

# Remove the duplicates from heart rate data

heartrate_seconds_cln_v4 <- heartrate_seconds_cln_v3 %>% 
  distinct()

# Duplicates observed and thus needs to be removed
mets_minute_cln_v3 %>% 
  get_dupes() %>% 
  summarise(count(id))

# Remove duplicates from METs data

mets_minute_cln_v4 <- mets_minute_cln_v3 %>% 
  distinct() 

# Duplicates observed in the data frame below

sleep_daily_cln_v2 %>% 
  get_dupes() %>% 
  summarise(count(id))

# Remove duplicates from sleep data

sleep_daily_cln_v3 <- sleep_daily_cln_v2 %>% 
  distinct()

# Final step of the cleaning process is to summarize the min and sec data into daily data in order to use for the daily insights.

heartrate_seconds_cln_dt <- data.table(heartrate_seconds_cln_v4)
heartrate_seconds_cln_dt_2 <- heartrate_seconds_cln_dt[,list(mean_hrt = mean(value), max_hrt = max(value), min_hrt = min(value)), by = c("id,date")]
  
  
mets_minute_cln_dt <- data.table(mets_minute_cln_v4)
mets_minute_cln_dt_2 <- mets_minute_cln_dt[,list(mean_mets = mean(mets), max_mets = max(mets), min_mets = min(mets)), by = c("id,date")]


# Using the summary function you can quickly identify if there are unusual numbers  
heartrate_seconds_cln_dt_2 %>% 
  summary()

mets_minute_cln_dt_2 %>% 
  summary()

activity_daily_cln_v2 %>% 
  summary()

sleep_daily_cln_v3 %>% 
  summary()

```

With the cleaning process completed and the data was ready for analysis. 

## Analyzing the data for key insights 

Descriptive statistic of the Daily Activity and Sleep Data: 

```{r, warning=FALSE}
# Analyze phase of the analysis 

# Starting with descriptive statistics for daily activity data

activity_daily_cln_dt <- data.table(activity_daily_cln_v2)
  
activity_daily_cln_dt_2 <- activity_daily_cln_dt[,list(total_active_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes)), by = c("id,date")]

activity_daily_cln_dt_3 <- activity_daily_cln_dt %>% 
  inner_join(activity_daily_cln_dt_2)

activity_daily_cln_dt_3$weekday <- weekdays(activity_daily_cln_dt_2$date)

activity_daily_mets_cln_merge <- activity_daily_cln_dt_3 %>% 
  inner_join(mets_minute_cln_dt_2)

tibble(activity_daily_mets_cln_merge)

activity_daily_mets_cln_merge %>% 
  select(total_steps, total_distance, tracker_distance, logged_activities_distance, very_active_distance, moderately_active_distance, light_active_distance, sedentary_active_distance, very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes, calories, total_active_minutes) %>% 
  summary()

# Descriptive statistics for sleep data 

tibble(sleep_daily_cln_v3)

sleep_daily_cln_dt <- data.table(sleep_daily_cln_v3)

sleep_daily_cln_dt_2 <- sleep_daily_cln_dt[,list(mean_minutes_asleep = mean(total_minutes_asleep), mean_minutes_in_bed = mean(total_time_in_bed)), by = c("id")]

sleep_daily_cln_dt_3 <- sleep_daily_cln_dt_2[,list(mean_minutes_not_sleeping = mean_minutes_in_bed - mean_minutes_asleep, mean_hour_asleep = mean_minutes_asleep / 60, mean_hour_in_bed = mean_minutes_in_bed / 60), by = c("id")]

sleep_daily_cln_dt_4 <- sleep_daily_cln_dt_2 %>% 
  inner_join(sleep_daily_cln_dt_3)

# ID 1844505072 is suspicious as they spent 16 hour sleeping, which is unusual. As for ID 2320127002 spending 1 hour asleep also seems unusual. These two anomalies can be due to the users device being incorrectly configured. 

tibble(sleep_daily_cln_dt_4)

sleep_daily_cln_v3 %>% 
  select(total_sleep_records, total_minutes_asleep, total_time_in_bed) %>% 
  summary()

sleep_daily_cln_dt_4 %>% 
  select(mean_minutes_asleep, mean_minutes_in_bed, mean_minutes_not_sleeping, mean_hour_asleep, mean_hour_in_bed) %>% 
  summary()

# Correlations found a negative correlation between sedentary min and calories burned which is expected, although the strong correlation between very active min and calories used is an interesting observation

correlation_activity <- activity_daily_mets_cln_merge %>% 
summarise(cor(total_active_minutes,calories), cor(very_active_minutes,calories), cor(fairly_active_minutes,calories), cor(lightly_active_minutes,calories), cor(sedentary_minutes,calories), cor(mean_mets,calories), cor(max_mets,calories), cor(min_mets, calories))

weekday_total_act_min <- activity_daily_cln_dt_3[,list(weekday_mean_total_act_min = mean(total_active_minutes), weekday_max_total_act_min = max(total_active_minutes), weekday_min_total_act_min = min(total_active_minutes)), by = c("weekday")] %>% 
  arrange(weekday_mean_total_act_min)

print(weekday_act_min_summary_stats)

```

The observed trends or relationships was as follow: 

	1. Very high activity minutes and calories used was strongly correlated
	2. Sedentary minutes and calories burned was negatively correlated
	3. The data for the Heart Rate and Sleep was incomplete which shows user's use this function on occasion. 
	4. Saturday was the most active day of the week
	5. From the data it was established that activity tracking is the primary use, therefore both the 'Bellabeats Leaf' and 'Bellabeats Time' would be sufficient.
	
## Visualizing the data insights

Visualizing the most active weekdays illustrated the specific days of the week where the activity levels where generally higher. Saturday was the most active day with Sunday being the least active. Traditionally, people work from Monday to Friday and have more free time on Saturday and Sunday.Most people tend to be more active on Saturdays and rest on Sundays before the next work week start.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Plotting the total activity minutes by weekday.

ggplot(weekday_act_min_summary_stats, aes(x = factor(weekday,c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")), y = round(weekday_mean_total_act_min, digits = 0), fill = weekday, label = round(weekday_mean_total_act_min, digits = 0))) + 
  geom_bar(stat = "identity") +
  labs(title = "Total Activity by day of the week", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "", y = "Total Activity Minutes", color = "black") + 
  geom_text(color = "black", vjust = 1.5) + 
  scale_fill_manual(values = Blue2Orange10Steps)

```

There is a strong correlation between very active minutes and calories  used in a day. As seen form the graph below:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Plotting correlations between very active minutes and calories.

ggplot(activity_daily_mets_cln_merge, aes(x = very_active_minutes, y = calories, color = calories)) +
  geom_point() +
  geom_smooth(method = lm, level = 0.99) +
  labs(title = "Very Active Minutes vs Calories", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "Very Active Minutes", y = "Calories")

```

The total active minutes and calorie usage is also correlated although it is interesting to note that the correlation is not as strong as with very active minutes. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Total active minutes comparison to calories.

ggplot(activity_daily_mets_cln_merge, aes(x = total_active_minutes, y = calories, color = calories)) +
  geom_point() +
  geom_smooth(method = lm, level = 0.99) +
  labs(title = "Total Active Minutes vs Calories", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "Total Active Minutes", y = "Calories")

```

FitBit makes use of METs which is a metric of activity minutes based on the graph below. It illustrates that their algorithm rewards more METs point to more vigorous activities and thus the correlation between METs is a strong compared to the total active minutes as seen below:

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Mean METs correlation to calories burned

ggplot(activity_daily_mets_cln_merge, aes(x = mean_mets, y = calories, color = calories)) +
  geom_point() +
  geom_smooth(method = lm, level = 0.99) +
  labs(title = "Mean METs vs Calories", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "Mean METs", y = "Calories")

```

Total step shows a strong correlation against calories burned during the day which is intuitive as the more you walk the more calories you will burn.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Plotting total step vs. calories

ggplot(activity_daily_mets_cln_merge, aes(x = total_steps, y = calories, color = total_steps)) +
  geom_point() +
  geom_smooth(method = lm, level = 0.99) +
  labs(title = "Total Steps vs Calories", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "Total Steps", y = "Calories", color = "Total Daily Steps")
```

The group slept less than the recommended 7 to 9 hour as prescribe by the CDC for adults with a mean of 6.3 hours of sleep per day. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Plotting sleep data 

ggplot(sleep_daily_cln_dt_4, aes(x = substr(id, 1, 3), y = mean_hour_asleep, fill = mean_hour_asleep, width = 0.5)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45)) +
  geom_hline(yintercept = c(7,9), size = 1.5) +
  geom_hline(yintercept = mean(sleep_daily_cln_dt_4$mean_hour_asleep), color = "green", size = 2, label = round(mean(sleep_daily_cln_dt_4$mean_hour_asleep), digits = 1)) +
  labs(title = "Mean Sleep Data", subtitle = "Data from 12-Apr-16 to 12-May-16", x = "Short ID", y = "Hours of Sleep", fill = "Mean Hours Asleep") +
  geom_label(aes(0, round(mean(sleep_daily_cln_dt_4$mean_hour_asleep),digits = 1), label = round(mean(sleep_daily_cln_dt_4$mean_hour_asleep), digits = 1), hjust = 0, vjust = 1.25)) +
  scale_y_continuous(breaks = c(1,2,3,4,5,6,7,8,9,10,11))
```

### Recommendation based on the Analysis

The CDC recommends 22 minutes per day of moderate activity the group which was analyzed had a mean of 13.6 fairly active minutes (moderate activity) per day, although combined with the very active minutes of 21.6 minutes the group analysed achieved there goal of 22 minutes as set out by the CDC.

The limited sleep data provides an opportunity for the app developers. In order to improve the 'sleep tracking function' suggestions can be offered to the users when they are getting too little sleep compared to the 7 to 9 hours of sleep which is recommended by the CDC. Additionally, a follow-up survey can be provided to the participants in an attempt to understand why they did not constantly sleep with their fitness device.

Based on the analysis the fitness devices used during the collection of the data was primaryily used for daily activity tracking and the two products that Bellabeats marketing team can focus on are the 'Bellabeats Leaf' and 'Bellabbeats Time'. Based on my experience with fitness devices one of the best features personalizing your fitness goals with your device. With that being said the app can focus on rewarding vigorous activities especially if the produce is being used for weight loss. Similarly Bellabeats can implement an activity metric such as METs or activity minutes which aligns with the CDC recommendations. Rewarding the user once they have achieved the minimum activity of 22 minutes and increasing the reward points as they increase pass the 22 minutes mark. 